---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# model formula
model_formula = bf(Distance~Sem*Phon) 

# checking priors
get_prior(model_formula, sim, family = "gaussian") 

# Specifying priors
my_priors <- c(prior(normal(107, 50), class = 'Intercept'),
                prior(normal(0,50), class = 'b'))

# Contrast-coding predictors
contrasts(sim$Sem) <-c(1,-1) 
contrasts(sim$Phon) <-c(1,-1)

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

# running model
model_sim <- brm(model_formula,
                  family = gaussian(),
                  prior = my_priors,
                  data = sim,
                  chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1)

```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

# Posterior predictive check
pp_check(model_sim, ndraws = 100)

# There looks to be somewhat of a mismatch between the real data and the simulated data, whereas the real data is right-skewed. A solution to the problem is to try a square root transformation of the Distance measure.


# Square root transformation
sim <- sim |> 
  mutate(sqrt_dist = sqrt(Distance))


# New model
model_sim2 <- brm(sqrt_dist~Sem*Phon,
                  family = gaussian(),
                  prior = my_priors,
                  data = sim)
                 

# Checking new model
pp_check(model_sim2, ndraws = 100)

#Looks much better!
```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# Output (new model with sqrt transformation)
model_sim2

#ASSESSMENT:
#Rhat values are ~ 1.00 = reasonably mixed
#Tail_ESS: > 100 = sufficient
#Bulk_ESS > 400 = sufficient

```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

# Visualizing posterior distributions:

#Checking variables
get_variables(model_sim2)

#Extract posteriors 
model_sim2 |> 
  spread_draws(b_Intercept,b_Sem1, b_Phon1,`b_Sem1:Phon1`)


#Plot Intercept 
library(tidybayes)
library(gridExtra)

plot1 <- model_sim2 |> 
  spread_draws(b_Intercept) |> 
  ggplot(aes(x = b_Intercept)) +
  stat_halfeye(fill = 'peachpuff') +
     labs(x = "Posterior Intercept",
        y = "")+
theme_minimal()
#Plot Sem1 
plot2 <-model_sim2 |> 
  spread_draws(b_Sem1) |> 
  ggplot(aes(x =b_Sem1)) +
  stat_halfeye(fill = 'darkorchid3') +
     geom_vline(xintercept = 0, lty = 'dashed') +
       labs(x = "Posterior Sem1",
        y = "")+
  theme_minimal()
#Plot Phon1 
plot3 <-model_sim2 |> 
  spread_draws(b_Phon1) |> 
  ggplot(aes(x =b_Phon1)) +
  stat_halfeye(fill = 'darkorchid3') +
     geom_vline(xintercept = 0, lty = 'dashed') +
       labs(x = "Posterior Phon1",
        y = "")+
  theme_minimal()
#Plot Interaction term
plot4 <-model_sim2 |> 
  spread_draws(`b_Sem1:Phon1`) |> 
  ggplot(aes(x =`b_Sem1:Phon1`)) +
  stat_halfeye(fill = 'darkorchid3') +
     geom_vline(xintercept = 0, lty = 'dashed') +
       labs(x = "Posterior Sem1:Phon1",
        y = "")+
  theme_minimal()


#Combining all plots 
grid.arrange(plot1, plot2, plot3, plot4, ncol = 2) 


#INTERPRETATION OF MODEL:
#Intercept: The posterior distribution of the model indicate that the average distance(square root transformed) across all combinations of categories is 8.47 [95% Credible Interval: 8.22 to 8.71]. 
#Sem1: The estimated average difference between semantically different and similar drawings is 0.34 [95% Credible Interval:0.9 to 0.59].
#Phon1: The estimated average difference between phonological different and similar drawings is -0.22 [95% Credible Interval:-0.47 to 0.02].
#Sem1:Phon1: The distribution for the average difference of differences(interaction term) is -0.05 [95% Credible Interval:-0.29 to 0.20].

```

```{r}
#| label: interpret_sim

# Calculating probability of direction (pd)
#Sem1:
model_sim2 |> 
  spread_draws(b_Sem1) |> 
  select(b_Sem1) |> 
  p_direction()
#Phon1:
model_sim2 |> 
  spread_draws(b_Phon1) |> 
  select(b_Phon1) |> 
  p_direction()
#Sem1:Phon1              
model_sim2 |> 
  spread_draws(`b_Sem1:Phon1`) |> 
  select(`b_Sem1:Phon1`) |> 
  p_direction()



# Calculating Bayes Factor
#extract posteriors
posteriors <- model_sim2 |> 
  spread_draws(b_Sem1, b_Phon1,`b_Sem1:Phon1`)|> 
  select(b_Sem1, b_Phon1,`b_Sem1:Phon1`)

#prior samples
model_sim2_priors <- brm(sqrt_dist~Sem*Phon,
                  family = gaussian(),
                  prior = my_priors,
                  sample_prior = "only",
                  data = sim)

priors <- model_sim2_priors |> 
    spread_draws(b_Sem1, b_Phon1,`b_Sem1:Phon1`)|> 
  select(b_Sem1, b_Phon1,`b_Sem1:Phon1`)

#Bayes factor against null value
bayesfactor_parameters(posteriors, priors, direction = "two-sided", null = 0)

```

## (g) Write up your entire analysis:

A Bayesian regression model was fitted to test the following hypothesis: (a) semantic similarity, (b) phonological similarity, and (c) their interaction affects how close/far apart participants draw cities on a map.

Posterior distributions of the model indicate that the average distance (square root transformed) across all combinations of categories is 8.47 \[95% Credible Interval: 8.22 to 8.71\]. The average difference in distance between semantically different and similar drawings is estimated to be an increase of 0.34 \[95% Credible Interval: 0.9 to 0.59\], and between the phonological different and similar to be decrease of -0.22 \[95% Credible Interval:-0.47 to 0.02\]. The posterior distribution for the interaction effect is estimated to an average difference of -0.05 \[95% Credible Interval:-0.29 to 0.20\].

The probability of direction (pd) was calculated for both predictors and their interaction. Given the current data, model and priors, 99.67% of the posterior samples for the semantic condition are positive, indicating that negative values are not very plausible and that the semantic condition is positively affecting the distance measure. For the phonological condition, 96.73% of the posterior samples are negative, indicating that positive values are not very plausible and that the phonological condition is negatively affecting the distance measure. For the posterior samples for the interaction effect, 64,3% of the samples are negative, indicating that both positive and negative values are plausible for this parameter.

The Savage-Dickey density ratio was subsequently calculated to assess evidence for and against the null hypothesis (H0= no difference in how close/far apart cities are drawn) using Bayes factor. The analysis yielded a Bayes Factor of 0.12 for the semantic condition, 0.012 for the phonological condition and 0.003 for the interaction, indicating that the posteriors relative to the priors have shifted from the point null value to a small degree. Given the data, the model and the priors, the alternative hypothesis that beta is not 0 is 12 times more probable than the null hypothesis (beta = 0) for the semantic condition, and 1,2 times more probable for the phonological condition. This is suggesting very small differences and small evidence for the alternative hypothesis.

------------------------------------------------------------------------

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
